---
title: "Portfolio Project - NASA APOD"
---

## Data Description

The NASA APOD dataset contains daily astronomy images and videos from 1995 to the present. It includes date, title, explanation text, image, URL, media type (image/video), and astronomer credits. This is the link: https://github.com/rfordatascience/tidytuesday/blob/main/data/2026/2026-01-20/readme.md. Some variables that I will be using include copyright (character), date, explanation (character), media_type (character), title (character), url (character), hdurl (character).

The research questions I will be trying to answer are: 

*RQ1* -  Which months produce the most images of galaxies vs. nebulae vs solar system objects? 

Answer: Solar system objects are consistently the most photographed object with July, January and March being the months that produce the most posts. See visualization 1 for more.

*RQ2* - Do explanation lengths correlate with image complexity? 

Answer: There is a weak positive correlation between explanation lengths and image complexity. See visualization 2 for more.

## Data Cleaning

For my data cleaning, I kept only the information I needed for my research questions and removed extra fields like URLs, HD URLS, and copyright information. I also cleaned up the media type variable so that all entries were consistently labeled as image, video, or other. To prepare for my visualizations, I created new variables from the text and date fields. One variable identifies the object type mentioned in each APOD entry, such as galaxy, nebula, or solar system object, by scanning the title and explanation for keywords. I also created a month and year variable from the date so I could look at patterns over time. For my second research question, I made a word-count variable to measure explanation length and a complexity variable based on how many different object types were mentioned. I also checked for missing or unclear values and planned to exclude or recode anything that did not fit clearly into my categories.

```{r, include = FALSE, message = FALSE}
#| label: Data-Cleaning
library(tidyverse)
library(lubridate)
library(stringr)

apod_raw <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-01-20/apod.csv")

#Transformation #1: Selecting relevant columns and dates, dropping url/hdurl/copyright as planned, converting date to proper date format for time-based grouping

apod_base <- apod_raw |>
  select(date, title, explanation, media_type) |>
  mutate(date = ymd(date)) 

#Transformation 2: Cleaning media_type and adding basic date feature, extracting month/year as ordered factors for RQ1 and RQ3 plotting

apod_timed <- apod_base |>
  mutate(
    media_type_clean = str_to_lower(str_trim(media_type)), #stringr for consistent casing
    media_type_clean = case_when(
      media_type_clean == "image" ~ "image",
      media_type_clean == "video" ~ "video",
      TRUE ~ "other" #Recode irregular types
    ),
    month = month(date, label = TRUE), #Ordered month for calendar
    year = year(date)
  )

#Transformation 3: Add RQ-specific text-derived variables, creating object_type by keyword matching (galaxy, nebula, solar, other), explanation_length for RQ2 and complexity as count of distinct objects mentioned

apod_final <- apod_timed |>
  mutate(
    text_content = str_to_lower(str_c(title, " ", explanation, sep = " ")), #Combine texts safely so no error pops up
    #Keyword matching for object_type
    has_galaxy = str_detect(text_content, "galaxy|milky way"),
    has_nebula = str_detect(text_content, "nebula|nebulae"),
    has_solar = str_detect(text_content, "planet|mars|jupiter|saturn|venus|mercury|earth|moon|sun"),
    object_type = case_when(
      has_galaxy & !has_nebula & !has_solar ~ "galaxy",
      has_nebula & !has_galaxy & !has_solar ~ "nebula",
      has_solar & !has_galaxy & !has_nebula ~ "solar",
      TRUE ~ "other" #For amiguous/multi-object cases
    ),
    explanation_length = str_count(explanation, boundary("word")), #Word count for RQ2
    complexity = (has_galaxy + has_nebula + has_solar) #Distinct types
  ) |>
  select(-text_content, -media_type)

```

## Visualization 1

The first visualization displays the APOD object type, which is either a galaxy, nebulae or solar object and is separated by month. As shown in the visualization solar is the overall most photographed object by NASA, followed by galaxies and nebulae.
```{r, echo = FALSE}
#| label: Data-Viz-RQ1
apod_final |>
  count(month, object_type) |>
  filter(object_type != "other") |>
  ggplot(aes(x = month, y = n, fill = object_type)) +
  geom_col() +
  labs(title = "APOD Object Types by Month",
       x = "Month",
       y = "Count",
       fill = "Object Type") +
  theme_light() +
  scale_fill_brewer(type = "qual", palette = "Set2") 
```
> Monthly counts of APOD object types by category

## Visualization 2

The second visualization shows the relationship between explanation length and image complexity for APOD entries. Each point represents one APOD entry, with explanation length measured by word count and complexity measured by the number of different object keywords mentioned in the title and explanation (e.g., solar, nebula, galaxy). The plot shows whether longer explanations tend to go with more complex images.
```{r, echo = FALSE, message = FALSE, fig.height = 4}
#| label: Data-Viz-RQ2

ggplot(apod_final, aes(x = explanation_length, y = complexity)) +
  geom_point(alpha = 0.6, color = "darkseagreen4") +
  geom_smooth(method = "lm", se = TRUE, color = "coral3") +
  labs(
    title = "Explanation Length and Image Complexity",
    x = "Explanation Length (Word Count)",
    y = "Complexity (Number of Object Types Mentioned)"
  ) +
  theme_bw()
```
> Scatterplot of APOD explanation length and coded image complexity.
